---
title: "Modeling"
format: html
editor: 
  markdown: 
    wrap: 72
---

# Modeling

## Introduction

Start with a basic introduction (feel free to repeat some things from
the other file).

## Data Split

Then, split the data into a training (70% of the data) and test set (30%
of the data). Set a seed to make things reproducible.

```{r echo=FALSE}
library(tidymodels)
set.seed(11)
diabetes_convert <- diabetes_convert |>
  mutate(Diabetes_binary = factor(Diabetes_binary))
diabetes_split <- initial_split(diabetes_convert, prop = 0.7)
diabetes_train <- training(diabetes_split)
diabetes_test <- testing(diabetes_split)

```

The goal is to create models for predicting the Diabetes_binary variable
(using tidymodels). We’ll use log-loss as our metric to evaluate the
models. For both model types, use log-loss with 5 fold cross-validation
to select the best model from that family of models. You should set up
your own grid of tuning parameters for each model (even if it is just
the number of levels to look at).

We are using our split data to create our 5 cross validation folds.

```{r}

diabetes_CV_folds <- vfold_cv(diabetes_train, 5, strata = Diabetes_binary)

log_loss_metric <- metric_set(mn_log_loss)
```

Now we can create the `recipe` we will use for all of the models, where
`Diabetes_binary` is the response/outcome.

```{r}
diabetes_rec <-
  recipe(Diabetes_binary ~., data = diabetes_train) |>
  step_novel(all_nominal_predictors()) |>
  step_unknown(all_nominal_predictors()) |>
  step_zv(all_predictors()) |>
  step_normalize(all_numeric_predictors())
diabetes_rec
```

## Classification Tree

You should provide a thorough explanation of what a classification tree
model is. Then you should fit a classification tree with varying values
for the complexity parameter and choose the best model (based on 5 fold
CV on the training set). Include at least 5 predictors in this model.

Let's define the model and the engine, where we want to using a tuning
grid to find the optimal `cost_complexity`.

```{r}
tree_mod <- 
  decision_tree(
    cost_complexity = tune()
  ) |>
  set_engine("rpart") |>
  set_mode("classification")
```

The next chunk of code is the creation of our workflow, where we add the
recipe and the model.

```{r}
tree_wkf <- workflow() |>
  add_recipe(diabetes_rec) |>
  add_model(tree_mod)
```

```{r}
tree_grid <- tibble(
  cost_complexity = c(0.0001, 0.001, 0.01, 0.1)
)
```

```{r}
tree_tune <- tune_grid(
  tree_wkf,
  resamples = diabetes_CV_folds,
  grid = tree_grid,
  metrics = log_loss_metric,
  control = control_grid(save_pred = TRUE)
)
tree_tune
```

```{r}
tree_tune |>
  collect_metrics()
```

```{r}
best_tree <- select_best(tree_tune, metric = 'mn_log_loss')
best_tree
```

## Random Forest

You should provide a thorough explanation of what a random forest is and
why we might use it (be sure to relate this to a basic classification
tree). You should then fit a random forest model with varying values for
the mtry parameter and choose the best model (based on 5 fold CV on the
training set). Include at least 5 predictors in this model.

```{r}
rf_mod <- rand_forest(mtry=tune()) |>
  set_engine('ranger') |>
  set_mode('classification')
rf_wkf <- workflow() |>
  add_recipe(diabetes_rec) |>
  add_model(rf_mod)
```

```{r}
#rf_fit <- rf_wkf |>
 # tune_grid(resamples = diabetes_CV_folds,
  #          grid = 7,
  #          metrics = log_loss_metric)
```

## Final Model Selection

You should now have two best models (one for each model type above).
Compare both models on the test set and declare an overall winner!
